Embedded Scalars: {A}n Experimental Revisit

### 2014-05-27 text/paper/paper4.tex

@@

\paragraph{Target Conditions.} We present sequences of pictures
depicting a set of four identical central elements (e.g., letters),
which could be connected to surrounding elements (e.g., triangles), as
shown in Figures~\ref{fig:exseqAS} and XYZ. Initially, any potential
connections between central and surrounding elements were covered by
dark gray color (see Figure~\ref{fig:exseqAS1}). Sentences to be
judged were presented auditorily, and participants were asked to
uncover the picture until they felt able to give a truth value
judgment. Three options were available for participants at each step:
(i) judge the sentence as true, (ii) judge the sentence as false,
(iii) demand more information (unavailable when the picture was fully
uncovered). Trials ended when a truth-value judgement was
made. Importantly, each of the three potentially available readings
corresponded to a specific step in the uncovering process where the
truth value of that reading (and only of that reading) could be
assessed for the first time. We refer to this step in the sequence as
the {\it critical position} of a given reading. The critical position
and the corresponding truth-value judgment differed between \as- and
\es-sentences, as described presently. (This is, partly, because of
the logical dependencies between readings, and, partly, also to rule
out positional biases.)

@@

Forty native speakers of German took part in our study, none of which
had any prior exposure to logic or formal semantics. We excluded two
participant due to insufficient performance on controls ($\le 60\%$
correct answers).

@@

We fitted generalized linear models to the contingency table data in
Table~\ref{tab:CC-ContingencyTable}, using factors \textsc{Condition}
with levels \as and \es, \textsc{Accent} with levels \emph{neutral}
and \emph{accented}, and \textsc{Position} with levels 1 through 3 to
predict the dependent factor \textsc{Reading} with levels
\emph{literal}, \emph{local}, and \emph{other}. The latter levels
lumps \emph{global} and \emph{error} responses together, because
otherwise we would not have enough cell counts to apply log-linear
regression. We selected the ``best'' predictor of the data by a
gradient search over hierarchical models in terms of AICs, starting
from the saturated model. The best model takes main factors
\textsc{Response} and \textsc{Condition} into account, as well as the
interaction term \textsc{Response:Condition} ($\chi^2 = 17.19$,
$df=42$, $p = 0.99$, $AIC = 158.01$ (compared to $AIC=208.25$ of the
saturated model)). Crucially, factors \textsc{Position} and
\emph{Accent} are spurious in the best model. This suggests that
response patterns only depend on the type of the sentence, but that
there were no learning or fatigation effects and that accentuation did
not influence answer patterns significantly.

\begin{table}
  \centering
  \begin{tabular}{lllll}
    coefficient & estimate & std.~error & $z$-value & $\Pr(>|z|)$ \\
    \midrule
    \textsc{Intercept} & 1.679  &  0.134 &  12.499  & < 0.001 \\
    \textsc{Reading}.\emph{literal} & -0.219  &  0.281 & -0.781 &
    0.435 \\
    \textsc{Reading}.\emph{local} & -2.249 &   0.172 & -13.084 & < 0.001 \\
    \textsc{Condition}.\es & 0.592 &  0.159 & 3.732 &  < 0.001 \\
    \textsc{Reading}.\emph{literal}:\textsc{Condition}.\es & -0.0186 &
    0.328 & -0.057 &  0.95471 \\
    \textsc{Reading}.\emph{local}:\textsc{Condition}.\es & 1.041 &
    0.208 &  4.999 &  < 0.001
  \end{tabular}
  \caption{Coefficients of the ``best'' log-linear model for the count
    data in Table~\ref{tab:CC-ContingencyTable}.}
  \label{tab:Coefficients-CC}
\end{table}

Inspection of the coefficients of the ``best'' model suggests that the
distinction between levels \emph{local} and \emph{other} in factor
\textsc{Reading} might not be necessary. In order to test whether
there is support for the postulation of local readings, we therefore
compared the previous ``best'' predictor to a model that only differs
from the former in that it subsumes the level \emph{local} of factor
\textsc{Reading} under level \emph{other} as well
\citep[e.g.][Chapter~15]{Crawley2007:The-R-Book}. Model comparison
reveals that there is no significant improvement in explanatory power
($Pr(|\chi|) = 0.2689$) of the more complex model that includes the
level \emph{local} (df=30, residual deviance = 9.76) over the simpler
model without this level (df=32, residual deviance = 12.39,
AIC=156.63).

This latter analysis suggests that our data does not provide strong
evidence for the maintenance of belief in local readings in our
setting. But we need to be careful. So far we have not taken into
account the sequential structure of our task. It might be that local
readings are infrequent because subjects tend to exit early in the
sequence, rather than late. This is why we need preference-related
controls.

\paragraph{Preference-related controls.} Positional answers for our
preference-related control conditions are shown in
Figure~\ref{Fig:TF_judgments}.\fn{We removed 19 data points for the
  \ec-conditions because of a coding mistake that presented subjects
  with incongruent sentence-picture pairs. Including these and
  classifying them as \emph{errors} does not change the results of the
  reported analyses.}
%
\begin{figure}[]
\centering

@@

As expected, most answers classify as \lc-readings, fewer as
\ec-readings. Moreover, intonation does seem to have the expected
effect as well: as compared to neutral intonation, \lc-cueing
intonation ups the count for \lc-readings, while \ec-cueing ups the
count for \ec-readings. If we compare the number of \lc- or
\ec-readings for each intonation type across sequences, we see that
there are always more answers indicative of the relevant readings when
that reading can be judged later in the sequence.

There are only very few answers that appear to be random
mistakes. Some answers in the `error' category are plausibly
`spill-over' errors. But on top of that, a substantial number of
`error' answers occur on critical positions. In the \lc-sequence there
are 23 \emph{true} answers at step 6 where \emph{false} answers would
be taken as indicative of \ec-readings. In the \ec-sequence there are
28 \emph{false} answers at step 5 where \emph{true} answers would be
taken as indicative of \lc-readings. \todo{A general tendency to
  answer later in the sequence does explain this. We don't see a
  similar effect in the critical conditions because the literal
  reading is so dominant.} Finally, intonation patterns that cue
\ec-readings seem to have led to more mistakes than other intonation
patterns.

@@

>>>>

\usetikzlibrary{calc}
\usepackage{nicefrac}

@@

\paragraph{Target Conditions.} We present sequences of pictures
depicting a set of four identical central elements (e.g., letters),
which could be connected to surrounding elements (e.g., triangles), as
shown in Figures~\ref{fig:exseqAS}--\ref{fig:exec}. Initially,
any potential connections between central and surrounding elements
were covered by dark gray color (see
Figure~\ref{fig:exseqAS1}). Sentences to be judged were presented
auditorily, and participants were asked to uncover the picture until
they felt able to give a truth value judgment. Three options were
available for participants at each step: (i) judge the sentence as
true, (ii) judge the sentence as false, (iii) demand more information
(unavailable when the picture was fully uncovered). Trials ended when
a truth-value judgement was made. Importantly, each of the three
potentially available readings corresponded to a specific step in the
uncovering process where the truth value of that reading (and only of
that reading) could be assessed for the first time. We refer to this
step in the sequence as the {\it critical position} of a given
reading. The critical position and the corresponding truth-value
judgment differed between \as- and \es-sentences, as described
presently. (This is, partly, because of the logical dependencies
between readings, and, partly, also to rule out positional biases.)

@@

Forty native speakers of German took part in our study, none of which
had any prior exposure to logic or formal semantics. We excluded two
participants due to insufficient performance on controls ($\le 60\%$
correct answers).

@@

We fitted generalized linear models to the contingency table data in
Table~\ref{tab:CC-ContingencyTable}, using factors \textsc{Condition}
with levels \as and \es, \textsc{Accent} with levels \emph{neutral}
and \emph{accented}, and \textsc{Position} with levels 1 through 3 to
predict the dependent factor \textsc{Reading} with levels
\emph{literal}, \emph{local}, and \emph{other}. The latter lumps
\emph{global} and \emph{error} responses together, because otherwise
we would not have enough cell counts to apply log-linear
regression. We selected the ``best'' predictor of the data by a
gradient search over hierarchical models in terms of AICs, starting
from the saturated model. The best model takes main factors
\textsc{Response} and \textsc{Condition} into account, as well as the
interaction term \textsc{Response:Condition} ($\chi^2 = 17.19$,
$df=42$, $p = 0.99$, $AIC = 158.01$ (compared to $AIC=208.25$ of the
saturated model)). Crucially, factors \textsc{Position} and
\textsc{Accent} are spurious in the best model. This suggests that
response patterns only depend on the type of the sentence, but that
there were no learning or fatigation effects and that accentuation did
not influence answer patterns significantly.

\begin{table}
  \centering
  \begin{tabular}{lllll}
    coefficient & estimate & std.~error & $z$-value & $\Pr(>|z|)$ \\
    \midrule
    \textsc{Intercept} & 0.916  &  0.258 &  3.49  & < 0.001 \\
    \textsc{Reading}.\emph{literal} & 2.600 &   0.268 & 9.716 & < 0.001 \\
    \textsc{Reading}.\emph{local} & -0.310  &  0.397 & -0.781 &
    0.435 \\
    \textsc{Condition}.\es & 1.030 &  0.301 & 3.423 &  < 0.001 \\
    \textsc{Reading}.\emph{literal}:\textsc{Condition}.\es & -1.288 &
    0.319 & -4.036 &  < 0.001 \\
    \textsc{Reading}.\emph{local}:\textsc{Condition}.\es & -0.026 &
    0.463 &  -0.0579 &  0.955
  \end{tabular}
  \caption{Coefficients of the ``best'' log-linear model for the count
    data in Table~\ref{tab:CC-ContingencyTable}. Notice that factor
    levels \emph{local} do not cause a significant shift in counts,
    given reference level
    \emph{false}.}
  \label{tab:Coefficients-CC}
\end{table}

Inspection of the coefficients of the ``best'' model suggests that the
distinction between levels \emph{local} and \emph{other} in factor
\textsc{Reading} might not be necessary (see
Table~\ref{tab:Coefficients-CC}). In order to test whether there is
support for the postulation of local readings, we therefore compared
the previous ``best'' predictor to a model that only differs from the
former in that it subsumes the level \emph{local} of factor
\textsc{Reading} under level \emph{other} as well
\citep[e.g.][Chapter~15]{Crawley2007:The-R-Book}. Model comparison
reveals that there is no significant improvement in explanatory power
($Pr(|\chi|) = 0.2689$) of the more complex model that includes the
level \emph{local} (residual df=30, residual deviance = 10.497) over the simpler
model without this level (residual df=32, residual deviance = 13.124,
AIC=157.268).

This latter analysis suggests that our data does not provide strong
evidence for the maintenance of belief in local readings in our
setting. But we need to be careful. The regression analysis tells us
that the number of errors and the number of local readings were
similar in all conditions. But the distribution of answers in
Figure~\ref{fig:JudgmentsK2} shows that small error counts occur at
many different steps, whereas we do see a concentrated rise in answers
at the critical position for local readings. Also, so far we have not
taken into account the sequential structure of our task. It might be
that local readings are infrequent because subjects tend to exit early
in the sequence, rather than late. This is why we need
preference-related controls and a more fine-grained analysis
eventually.

\paragraph{Preference-related controls.} Positional answers for our
preference-related control conditions are shown in
Figure~\ref{Fig:TF_judgments}.\fn{We removed 19 data points for the
  \ec-conditions because of a coding mistake that presented subjects
  with incongruent sentence-picture pairs. Including these and
  classifying them as \emph{errors} does not change the qualitative
  results of the reported analyses.}
%
\begin{figure}[]
\centering

@@

As expected, most answers classify as \lc-readings, fewer as
\ec-readings. Moreover, intonation does seem to have the expected
effect as well: as compared to neutral intonation, \lc-cueing
intonation increases the count for \lc-readings, while \ec-cueing
increases the count for \ec-readings. If we compare the number of \lc-
or \ec-readings for each intonation type across sequences, we see that
there are always more answers indicative of the relevant readings when
that reading can be judged later in the sequence. This suggests a
general tendency to answer later in the sequence, rather than at the
earliest possible step.

There are only very few answers that appear to be random
mistakes. Some answers in the `error' category are plausibly
`spill-over errors', where participants gave answers one step too
late. But on top of that, a substantial number of `error' answers
occurs on critical positions. For example, in the \lc-sequence there
are 23 \emph{true} answers at step 6 where \emph{false} answers would
be taken as indicative of \ec-readings. In the \ec-sequence there are
28 \emph{false} answers at step 5 where \emph{true} answers would be
taken as indicative of \lc-readings.\fn{A general tendency to answer
  later in the sequence would explain this. The reason why we do not
  see a similar effect in the critical conditions, might be because
  the literal reading is so dominant.} Finally, intonation patterns
that cue \ec-readings seem to have led to more mistakes than other
intonation patterns.

@@

\subsubsection{Probabilistic salience competition.}

Our previous regression-based analyses did not take the sequential
nature of the task into account. We initially hypothesized that
subjects might exit as early as possible, but the data from the
target-related controls suggests otherwise. In order to estimate this
bias and the preference for candidate readings, we use a simple
Bayesian cognitive model that computes the likelihood of answer
patterns for different latent sequential biases and reading
preferences. We use the data to estimate the posterior likelihood of
these latent parameters to gain insight into the likely preference
relations and the salience of local readings.

\paragraph{The model.} We outline a simple probabilistic model that
quantifies the positional bias and the competing strength of available
readings. Take the AS-conditions as an example. There are three
potential readings. The relative strength between these is given by a
probability vector $\myvec{p}^\text{as-n} =
\tuple{p^\text{as-n}_{\text{lit}},p^\text{as-n}_{\text{loc}},p^\text{as-n}_{\text{glb}}}$. We
think of $p^\text{as-n}_{\text{lit}}$, for instance, as the salience
of the literal reading for \as-sentences with neutral prosody,
relative to the global and local reading. If there were no potential
effects of sequentiality and no errors (or other readings that we
classify as errors), the vector $\myvec{p}^\text{as-n}$ would be our
prediction of observed frequencies of answer patterns. We look at four
such three-placed vectors, one for each target condition.

For the preference-related control conditions we look at three vectors
$p^k = \tuple{p^{\text{k}}_{\text{\lc}}, p^{\text{k}}_{\text{\ec}}}$,
where $k \in \set{text{ntr}, \text{\lc-cue}, \text{\ec-cue}}$. These
vectors give the relative salience of \lc- and \ec-readings for
different types of intonation. The salience of readings is independent
of the pictorial sequence. If there is a tendency to judge sentences
earlier in the sequence, rather than later, that is to be attributed
to the sequential bias, not a difference in general, task-independent
salience of readings.

To add potential effects of sequentiality, we assume that there is a
constant bias factor $q \in [0;1]$. This bias is held constant over
all conditions, because we assume that there is a global tendency to
answer early or late, independent of the sentence subjects are
confronted with. The bias factor captures whether there is a tendency
to answer early ($q > .5$) or later ($q <.5$) in the sequence. In the
\as-sequences, the literal reading can be judged first, then the local
and then the global reading. So we will assume that $p_{\text{lit}}$
is multiplied by $q$, that $p_{\text{loc}}$ is multiplied by $(1-q)q$
and that $p_{\text{glb}}$ is multiplied by $(1-q)^2$.

Finally, we also allow for the chance of subjects making mistakes. To
keep things as simple as possible (no consideration for spill-overs,
late decisions, preferences to unravel completely before judging
etc.), we assume that there is a fixed error rate $e^k \in [0;1]$ for
each sentence-intonation pair $k$. We allow for different error
rates for different sentence-intonation pairs, because we want to
allow for the possibility that non-monotonic quantifiers might be
harder to process, or that some kind of intonation (like the \ec-cue)
is harder than others. A decision can be wrong but incidentally
coincide with a reading that is indicative of a relevant
reading. Error rates multiply in proportion to the number of steps in
a sequence: more choice, more chance to be wrong.

\bigskip

HERE

Taking this together we calculate, for each critical
sentence-intonation pair $k$ a target probability vector $\myvec{t}^k
=
\tuple{t^k_{\text{lit}},t^k_{\text{loc}},t^k_{\text{glb}},t^k_{\text{err}}}$
with which we expect an answer to be classified as LIT, LOC, GLB or as
an error. For example, for AS-N, where LIT can be judged before GLB
before LOC, we get:
\begin{align*}
  t^{\text{as-n}}_{\text{lit}} & \propto q \
  p^\text{as-n}_{\text{lit}} + e^\text{as-n} &   t^{\text{as-n}}_{\text{loc}} & \propto (1-q)^2 \
  p^\text{as-n}_{\text{loc}} + e^\text{as-n} \\
  t^{\text{as-n}}_{\text{glb}} & \propto (1-q)q \
  p^\text{as-n}_{\text{glb}} + e^\text{as-n} &   t^{\text{as-n}}_{\text{err}} &
  \propto 12 e^\text{as-n}
\end{align*}
Two remarks. The vector $\myvec{t}^k$ should be normalized eventually,
which is why $\propto$ is used instead of $=$. Also, the probability
of error $t^{\text{as-n}}_{\text{err}}$ comes from the observation
that there are 15 positions in total in an AS-sequence at which
subjects can make an erroneous choice, but we have accounted for three
of these already as adding to the count of answers indicative of
relevant readings.

Similarly, we compute a target probability $\myvec{t}^k =
\tuple{t^k_{\text{late}},t^k_{\text{early}},t^k_{\text{err}}}$ for
each preference-related control condition $k$, taking into account
which reading can be judged first and the length of the respective
sequences.

The full probabilistic model is visualized in
Figure~\ref{fig:model_graph}. As indicated there, we assume largely
uninformative priors. Any positional bias $q \in [0;1]$ is assumed to
be equally likely. The error rates for each condition should be small,
so we sample uniformly from interval $[0,.2]$. The biases
$\myvec{p}_k$ are also determined uniformly at random, by sampling from a
Dirichlet distribution with equal weights on all dimensions.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[node distance = 2.5cm, 
    double distance = 2pt,
    minimum size=1.25cm,
    thick]

    \node[rectangle, draw=black, fill=black!30] (obs)
    {obs$_k$};

    \node[rectangle, draw=black, fill=black!30,
    right of = obs] (n) {$n_k$};

    \node[below of = n, node distance= 1.5cm]
    (anchor) {};

    \node[left of = anchor, node distance= 0.5cm] (k) {$k \in
      \set{\text{as-n, as-e, \dots}}$};

    \node[circle, double, draw=black,
    above of = obs] (t) {$\myvec{t}_k$};

    \node[circle, draw=black, 
    above of = t] (p) {$\myvec{p}_k$};

    \node[circle, draw=black, 
    right of = p] (e) {$e_k$};

    \node[circle, draw=black, 
    left of = p] (q) {$q$};

    \draw[->] (t)--(obs);

    \draw[->] (n)--(obs);

    \draw[->] (q)--(t);

    \draw[->] (p)--(t);

    \draw[->] (e)--(t);

    \begin{pgfonlayer}{background}
       \node [rounded corners,
       very thick,draw=black!40,fit={($(obs.south)+(0,-30pt)$) 
                                 ($(n.east)+(+5pt,0)$) 
                                 ($(e.north)+(0,+5pt)$) 
                                 ($(p.west)+(-5pt,0)$)}] {};
    \end{pgfonlayer}

    \begin{scope}[xshift = 7.5cm, node distance = 1cm]

      \node[] (obsd) {$\text{obs}_k \sim \text{Multinomial}(\myvec{t}_k,n_k)$};

      \node[above of = obsd] (td) {$\myvec{t}_k = \text{as described
          in text}$};

      \node[above of = td] (pd) {$\myvec{p}_k =
        \text{Dirichlet}(\tuple{\nicefrac{1}{k}, \nicefrac{1}{k}, \dots})$};

      \node[above of = pd] (ed) {$e_k \sim \mathcal{U}(0,.2)$};

      \node[above of = ed] (qd) {$q \sim \mathcal{U}(0,1)$};

    \end{scope}

  \end{tikzpicture}
  \caption{Probabilistic graphical model}
  \label{fig:model_graph}
\end{figure}
